%%% Please fill in basic information on your thesis, which will be automatically
%%% inserted at the right places.

% Type of your thesis:
%	"bc" for Bachelor's
%	"mgr" for Master's
%	"phd" for PhD
%	"rig" for rigorosum
\def\ThesisType{bc}

% Language of your study programme:
%	"cs" for Czech
%	"en" for English
\def\StudyLanguage{cs}

% Thesis title in English (exactly as in the official assignment)
% (Note: \xxx is a "ToDo label" which makes the unfilled visible. Remove it.)
\def\ThesisTitle{Streamlining Usability of Enterprise Data Quality Management Tools for Data Engineers}

% Author of the thesis (you)
\def\ThesisAuthor{Zdeněk Tomis}

% Year when the thesis is submitted
\def\YearSubmitted{2024}

% Name of the department or institute, where the work was officially assigned
% (according to the Organizational Structure of MFF UK in English,
% see https://www.mff.cuni.cz/en/faculty/organizational-structure,
% or a full name of a department outside MFF)

% department assigned to the thesis
\def\Department{Department of Distributed and Dependable Systems}
% Is it a department (katedra), or an institute (ústav)?
\def\DeptType{Department}

\def\Supervisor{doc. Ing. Lubomír Bulej, Ph.D.}
\def\SupervisorsDepartment{Department of Distributed and Dependable Systems}

% Study programme and specialization
\def\StudyProgramme{Computer science}

% An optional dedication: you can thank whomever you wish (your supervisor,
% consultant, who provided you with tea and pizza, etc.)
\def\Dedication{%
I would like to dedicate this thesis to my family, and friends, who support has been and continues to be priceless. Special thanks go to my supervisor, doc. Ing. Lubomír Bulej, Ph.D., for his guidance, flexibility, and understanding. I would also like to express my gratitude to colleagues in Ataccama for their support, namely to Lukáš Kolek and Ondřej Chrastina.
}

% Abstract (recommended length around 80-200 words; this is not a copy of your thesis assignment!)
\def\Abstract{%
In the realm of data quality management, integrating robust data quality rules into automated workflows and data pipelines is essential for maintaining data integrity. This thesis addresses the gap in programmatic accessibility of Ataccama ONE’s data quality tools, which primarily leverage the proprietary Ataccama Expression Language. By reimplementing this language in Python, the project enhances its usability for data engineers who seek to consume these tools programmatically. The focus is on enabling data engineers to execute Ataccama’s rules directly within Python. The viability of this implementation is tested through performance comparisons with similar solutions.
}

% 3 to 5 keywords (recommended) separated by \sep
% Keywords are useful for indexing and searching for the theses by topic.
\def\ThesisKeywords{%
data quality management\sep data engineering\sep performance evaluation
}

% If any of your metadata strings contains TeX macros, you need to provide
% a plain-text version for use in XMP metadata embedded in the output PDF file.
% If you are not sure, check the generated thesis.xmpdata file.
\def\ThesisAuthorXMP{\ThesisAuthor}
\def\ThesisTitleXMP{\ThesisTitle}
\def\ThesisKeywordsXMP{\ThesisKeywords}
\def\AbstractXMP{\Abstract}

% If your abstracts are long and do not fit in the infopage, you can make the
% fonts a bit smaller by this setting. (Also, you should try to compress your abstract more.)
\def\InfoPageFont{}
%\def\InfoPageFont{\small}  % uncomment to decrease font size

% If you are studing in a Czech programme, you also need to provide metadata in Czech:
% (in English programmes, this is not used anywhere)

\def\ThesisTitleCS{Zjednodušení použitelnosti nástrojů pro správu kvality dat pro datové inženýry}
\def\DepartmentCS{Katedra distribuovaných a spolehlivých systémů}
\def\DeptTypeCS{Katedra}
\def\SupervisorsDepartmentCS{Katedra distribuovaných a spolehlivých systémů}
\def\StudyProgrammeCS{Informatika}

\def\ThesisKeywordsCS{%
data quality management\sep data engineering\sep performance evaluation
}

\def\AbstractCS{%
V oblasti data quality managementu je pro zachování integrity dat zásadní integrovat pravidla kvality dat do automatizovaných workflows a datových pipelines. Tato práce se zabývá mezerou v programové dostupnosti nástrojů pro kvalitu dat společnosti Ataccama ONE, které využívají především proprietární jazyk Ataccama Expression Language. Reimplementací tohoto jazyka v jazyce Python projekt zvyšuje jeho použitelnost pro datové inženýry, kteří potřebují tyto nástroje využít programmaticky v různým prostředích. Důraz je kladen na to, aby datoví inženýři mohli provádět a spravovat pravidla Ataccama přímo v jazyce Python s ohledem na jednoduchost užití a minimální nároky. Užitelnost této implementace je otestována prostřednictvím porovnání výkonu s podobnými řešeními.
}
